---
title: "human B cell atlas and deriving isotype signatures"
output: html_notebook
---
```{r setup}
library(Seurat)
library(sciCSR)
```

```{r}
# read Stewart et al & King et al atlas
pure <- 'scPure2_Pure1Integrated_noPC_IGHC_metacells.rds'
pure <- readRDS(pure)
pure <- subset(pure, subset = (orig.ident == "HB6")) # the original donor
DefaultAssay(pure) <- 'integrated'

king <- 'SEURAT_OBJECTS/HumanTonsil_BCells_scRNA_IGHC_metacell_SeuratObject.rds'
king <- readRDS(king)

pure <- NormalizeData(pure)
pure <- FindVariableFeatures(pure, selection.method = "vst", nfeatures = 2000)
king <- NormalizeData(king)
king <- FindVariableFeatures(king, selection.method = "vst", nfeatures = 2000)
```

```{r}
# integrate two atlas into one for fitting trajectory using slingshot
features <- SelectIntegrationFeatures(object.list = list(pure, king))
anchors <- FindIntegrationAnchors(object.list = list(pure, king), anchor.features = features)
combined <- IntegrateData(anchorset = anchors)
```

```{r}
# Run the standard workflow for visualization and clustering
combined <- ScaleData(combined, verbose = FALSE)
combined <- RunPCA(combined, npcs = 30, verbose = FALSE)
combined <- RunUMAP(combined, reduction = "pca", dims = 1:30)
combined <- FindNeighbors(combined, reduction = "pca", dims = 1:30)
combined <- FindClusters(combined, resolution = 0.5)

DimPlot(combined, group.by = "orig.ident")
DimPlot(combined, group.by = "transfer_label", label = TRUE)
DimPlot(combined, group.by = "CellType", label = TRUE)
DimPlot(combined, label = TRUE)
```

```{r}
saveRDS(combined, '/media/josefng/Seagate4TB/GLT_datasets/scPure-HB6_KingTonsil_integrated.rds')
```

One easy way to derive a score would be to decompose the productive/sterile count matrix for "isotype signatures" - one would expect a prominent signature being a naive like signature (i.e. IgM-dominated) - and then extract the weight (coefficient) associated with this signature for each cell. And then invert this such that naive cells have a small number. The signature matrix can be stored and decompose future samples (i.e. new datasets) using nonnegative least square (NNLS) regression.

The matrix decomposition was performed using nonnegative matrix factorization (NMF). Here a simple NMF decomposition into 2 signatures (i.e. rank = 2) were performed - the idea is oen should expect a naive signature (heavy weights for IgM productive/sterile) and a memory signature. The safest seems to be to use the naive signature (as memory can be biased towards any isotype depending on the datasets, both reference data (i.e. our atlases)) or new datasets from the user).

```{r}
combined <- readRDS('scPure-HB6_KingTonsil_integrated.rds')
```

```{r}
stewart <- subset(combined, subset = (orig.ident == "HB6"))
king <- subset(combined, subset = (Lineage == "B Cells"))
library(NMF)
normcounts <- list("stewart" = as.matrix(stewart@assays$IGHC@data),
                   "king" = as.matrix(king@assays$IGHC@data))
normcounts[["stewart"]] <- normcounts[["stewart"]][!grepl("^IGHD", rownames(normcounts[["stewart"]])), ]
normcounts[["king"]] <- normcounts[["king"]][!grepl("^IGHD", rownames(normcounts[["king"]])), ]
normcounts[["stewart"]] <- normcounts[["stewart"]][!grepl("-C$", rownames(normcounts[["stewart"]])), ]
normcounts[["king"]] <- normcounts[["king"]][!grepl("-C$", rownames(normcounts[["king"]])), ]
normcounts[["stewart"]] <- normcounts[["stewart"]][, colSums(normcounts[["stewart"]]) > 0]
normcounts[["king"]] <- normcounts[["king"]][, colSums(normcounts[["king"]]) > 0]
saveRDS(normcounts, "scPure-HB6_KingTonsil_IGHC_NormCounts.rds")
```

```{r}
normcounts <- readRDS("scPure-HB6_KingTonsil_IGHC_NormCounts.rds")
normcounts <- cbind(normcounts$stewart, normcounts$king)
```

```{r}
# test NMF of different ranks. set upper bound = 8 as there are 8 isotypes except IgD
library(NMF)
test_nmf <- list()
test_nmf <-  nmf(normcounts, rank = 2:8, method = "snmf/r", 
                 nrun = 10, .options = "p4 v")
```


```{r}
# final rank=2 nmf
nmf <- nmf(normcounts, rank = 2, method = "snmf/r", nrun = 10, .options = "p2 v")
saveRDS(nmf, "scPure-HB6_KingTonsil_IGHC_NMF-combined.rds")
```

## Mouse

```{r}
mouse <- readRDS("mouse_Bcells_atlas_IGHC.rds")
normcounts <- as.matrix(mouse@assays$IGHC@data)
normcounts <- normcounts[!grepl("^Ighd", rownames(normcounts)), ]
normcounts <- normcounts[!grepl("-C$", rownames(normcounts)), ]
normcounts <- normcounts[, colSums(normcounts) > 0]
saveRDS(normcounts, "mouse_Bcells_atlas_IGHC_NormCounts.rds")

# run nmf
library(NMF)
set.seed(1234)
test <- normcounts[,sample(1:ncol(normcounts), 40000)]
nmf <- nmf(test, rank = 2, method = "snmf/r", nrun = 10, .options = "p2 v")
saveRDS(nmf, "scPure-HB6_KingTonsil_IGHC_NMF-combined.rds")
```

```{r}
# this is the signature matrix
pheatmap::pheatmap( nmf@fit@W, cluster_rows = FALSE, cluster_cols = FALSE )

# the first column is the naive signature

# fit the coefficients of these two signature for each cell in the data
# using nnls

naive_signature <- sapply(Cells(combined), 
                          function(n) nnls::nnls(nmf@fit@W, combined@assays$IGHC@data[rownames(nmf@fit@W), n])$x[1])
# scale into range [0, 1] and invert
naive_signature <- (naive_signature - min(naive_signature)) / abs(diff(range(naive_signature)))
naive_signature <- 1 - naive_signature
combined <- AddMetaData(combined, naive_signature, col.name = "csr_score")

ggplot(combined@meta.data, aes(x = csr_pot, y = csr_score)) + geom_point()

FeaturePlot(combined, c("csr_pot", "csr_score"))
```

